---
title: "Estimating Vaccine Efficacy: A Comparative Analysis of Method of Moments and Maximum Likelihood Estimation"
subtitle: "Authors: Katherine Fulton, Kreslyn Hinds, Tracia Pan"
graphics: yes
output: 
        pdf_document:
         toc: false
         number_sections: true
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{multicol,graphicx,hyperref,xcolor}
- \usepackage{setspace} \doublespacing
fontsize: 11pt

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(maxLik)
library(patchwork)
```

# Abstract
**Introduction** 

This study estimates the efficacy of Pfizer's (BNT162b2) vaccination for COVID 19, and compares the estimate with the Food and Drug Administration's Emergency Use Authorization requirement of 30% efficacy.

**Statistical Methods**

Maximum likelihood estimation (MLE), method of moments (MOM), and bootstrap methods were applied to estimate the efficacy of the vaccination. Confidence intervals were constructed and a likelihood ratio test was conducted to test the null hypothesis.

**Results**

Both MLE and MOM estimated an vaccine efficacy wee above 30%. In addition, both the likelihood ratio test and the empirical P-value hypothesis test resulted in very small values.

**Discussion**

Both confidence intervals had a lower bound around 90%, as 30% is not in either confidence interval we conclude that the efficacy rate was significantly higher than 30%. This conclusion is further supported by both p-values calculated in our hypothesis tests being effectively 0.
     
## Keywords
* Maximum Likelihood Estimation: Using the likelihood function of a parameter, the maximum likelihood estimation approach for estimating a parameter finds the maximum of the function after a log transformation by using it's first and second derivative. After finding the maximum of the function we can solve for the estimated parameter $^{6.1}$.
* Method of Moments Estimation: With the assumption that the expected value of a random variable will be a function of a parameter and the sample average will be close to the expected value, methods of moments estimation sets the sample average equal to the expected value and solves for the estimated parameter $^{6.2}$.
* Food and Drug Administration (FDA) Emergency Use Authorization (EUA): A mechanism of the FDA to enable the usage of medical countermeasures during public health emergencies. This may include allowing use of unapproved medical products or allowing the usage of already approved medical products for an unapproved purpose $^5$. 
* Coronavirus (COVID 19): A contagious infectious disease caused by SARS-CoV-2, first originating in Wuhan, China December 2019 $^1$.
* Pfizer Vaccine (BNT162b2):A two-dose vaccine for SARS-CoV-2, safe for all individuals 6 months and above $^3$.

## Introduction
With a total of 775 million cases and 7 million deaths reported from January 2020-May 2024 $^4$, Coronavirus has affected people around the globe and continues to be an ongoing concern. After being declared a pandemic by the World Health Organization on March 11, 2020 countries scrambled to mount a defense against this deadly virus. Reactive measures in the United States included travel bans and statewide lock downs, while hospitals filled with patients$^1$. To reach a more stable solution an effective vaccine was needed. Once companies started developing this lifesaving technology, emergency use authorizations were issued to accelerate the approval of these vaccines and allow them to be used. 

In December 2020, Pfizer and BioNTech successfully obtained a U.S. FDA Emergency Use Authorization (EUA) to begin distributing their vaccine (BNT162b2) for COVID 19 $^{6.3}$. To obtain the emergency use authorization, an efficacy rate of at least 30% is required $^{6.4}$. Using data as reported by Polack et al.$^2$, we will use maximum likelihood estimation and method of moments estimation to report the estimated efficacy rate of the Pfizer vaccine. Using these estimates we will create confidence intervals for the efficasy estimate and preform hypothesis testing to determine if the efficacy rate meets the EUA requirements of at least 30% $^{6.4}$. The data we're using is collected from a placebo-controlled, observer-blinded efficacy study, which randomly assigned persons 16 years of age or older in a 1:1 ratio to receive two doses, 21 days apart, of either placebo or the candidate, BNT162b2 vaccine. The primary efficacy endpoint was the time of laboratory-confirmed COVID-19. A total of 170 infections were targeted$^{6.3}$.

We will specifically focus on the relevant data from Table 2 of the paper referenced above (summarized below):
\begin{table}[h]
\centering
\caption{Vaccine Efficacy against Covid 19 at least 7 days after second dose in patients without evidence of infection}
\begin{tabular}{cccc}
Group & Covid-19 Cases & No. of subjects\\ \hline
BNT162b2 & 8 & 17,411 \\
Placebo & 162 & 17,511 \\ \hline
Total & 170 & 34,922 \\ \hline
\end{tabular}
\end{table} 

```{r, echo = FALSE, include = FALSE}
# create data frame from table
Vaccine_data <- data.frame(
  Group = c("BNT162b2", "Placebo"),
  Covid19_Cases = c(8, 162),
  No_of_Subjects = c(17411, 17511)
)

# calculate the number of non-cases
Vaccine_data <- Vaccine_data %>%
  mutate(Non_Covid19_Cases = No_of_Subjects - Covid19_Cases)

print(Vaccine_data)

# observed vaccine efficacy
vac_prob <- 8 / 17411
plac_prob <- 162 / 17511

vac_eff <- (1 - (vac_prob / plac_prob))* 100
```   

# Statistical Methods

Using a single binomial variable model, let $T$ denote the number of subjects in the BNT162b2 group out of all the $n = 170$ Covid-19 cases.
\begin{align}
\begin{aligned}
T \sim (n = 170, \pi) \\
\text{Where } \pi = P(BNT162b2|Covid-19 case) \text{ and can be shown to be equal to} \\
\pi = \frac{n_1 \pi_1}{n_1 \pi_1+n_2 \pi_2}\\
\text{Since } n_1 = 17411 \text{ and } n_2 = 17511\text{, we can say that }\\ 
n_1 \approx n_2 \text{, the randomization is } 1:1 \text{, and can simplify } \pi \text{ to} \\
\pi = \frac{\pi_1}{\pi_1+ \pi_2} \\
\text{We then have our parameter of interest} \\
\psi = \frac{\pi_2 - \pi_1}{\pi_2} \\
= \frac{1 - 2\pi}{1-\pi} \\
= 1 - \frac{\pi_1}{\pi_2}
\end{aligned}
\end{align}

To make inferences, we decided to use likelihood inference and method of moments estimation.

## Likelihood inference
In this section, the $\hat{\psi_0^{mle}}$ was calculated, and large sample confidence interval and a likelihood ratio test were performed to calculate the P-value that tests $H_0: \psi_0 = 0.3$ versus $H_a: \psi_0 > 0.3$. $\\$
parameter of interest: $\psi = \frac{1 - 2\pi}{1-\pi} \Longleftrightarrow \pi=\frac{1- \psi}{2- \psi}\\$  
\begin{align}
\begin{aligned}
L(\psi) &=  (\binom{170}{t}) (\frac{1- \psi}{2- \psi})^t (1-\frac{1- \psi}{2- \psi})^{170-t}, x = 0, 1, \dots, 170 \\
\ell(\psi) &= ln(\binom{170}{t}) + tln(\frac{1- \psi}{2- \psi}) + (170-t)ln(1-\frac{1- \psi}{2- \psi})\\
\ell '(\psi) &= -\frac{t}{(-\psi+1)(-\psi+2)} + \frac{170-t}{\psi + 2}\\
\end{aligned}
\end{align}

\begin{align}
\begin{aligned}
\text{To find } \hat{\psi_0^{mle}} \text{, we set } \ell '(\psi) &= 0\\
-\frac{t}{(-\psi+1)(-\psi+2)} + \frac{170-t}{\psi + 2} &= 0\\
\frac{t}{(-\psi+1)(-\psi+2)} &= \frac{170-t}{\psi + 2} \\
t(-\psi+2) &= (170-t)(-\psi+1)(-\psi+2)\\
t &= (170-t)(-\psi+1)\\
t &= -170\psi + \psi t - t +170 \\
170 \psi - \psi t &= -2t+170\\
\psi(170-t) &= -2t+170 \\
\hat{\psi_0^{mle}} &= \frac{-2t+170}{170-t}\\
t &= 8 \\
\hat{\psi_0^{mle}} &= \frac{-2(8)+170}{170-(8)} = 0.95
\end{aligned}
\end{align}
We also used the maxLik to find the Maximum Likelihood estimation (See Appendix 1 for the r code that was used) $\\$

To find the 95% confidence interval, we need to find $SE = \sqrt{\frac{-1}{\ell ''(\psi)}} \\$
\begin{align}
\begin{aligned}
\ell '(\psi) &= -\frac{t}{(-\psi+1)(-\psi+2)} + \frac{170-t}{\psi + 2}\\
\ell ''(\psi) &= \frac{t(2\psi-3)}{(-\psi +1)^2(-\psi+2)^2} + \frac{170-t}{(-\psi+2)^2}\\
\text{Plugging in } t = 8 \text{ and } \psi = 0.95\\
&= -3045.805 \\
- \ell ''(\psi) &= 3045.805\\
SE &= \sqrt{\frac{-1}{\ell ''(\psi)}} = \sqrt{\frac{1}{-\ell ''(\psi)}} \\
&= \sqrt{\frac{1}{3045.805}} \\ 
\end{aligned}
\end{align}

Assuming that n is large and the regularity conditions are met, we have $\\$
$\psi_0^{mle} \approx Norm(\psi_0, SE(\psi_0^{mle})= \sqrt{\frac{1}{3045.805}})\\$
The 95% Confidence interval is $\\$
$\psi_0^{mle} \pm 1.96 \times SE \\$
$0.95 \pm 1.96 \times \sqrt{\frac{1}{3045.805}} \\$
(See Appendix 2 for the r code used for calculation) $\\$

Using the Likelihood ratio test to find the p-value, we have the hypothesis $\\$
$H_0: \psi_0 = 0.3$ $H_a: \psi_0 > 0.3$. $\\$
$\hat{\psi_0^{mle}} = 0.95\\$
$W = 2ln(\frac{L(\psi_0^{mle})}{L(\psi_0^{mom})}) \approx 121.6\\$
When $H_0$ is true, $W \approx \chi_1^2\\$
P-value $= P(\chi_1^2 \geq 121.6) \\$
(See Appendix 3 for the r code used to calculate the P-value) $\\$

## Method of moments estimation
In this section, we found the $\hat{\psi_0^{mom}}$, confidence intervals for $\psi_0$ using bootstrap, and the empirical P-values to calculate the P-value that tests $H_0: \psi_0 = 0.3$ versus $H_a: \psi_0 > 0.3$. $\\$

We observed $t = 8$ and the distribution is $T \sim (n = 170, \pi) \\$
parameter of interest: $\psi = \frac{1 - 2\pi}{1-\pi} \Longleftrightarrow \pi=\frac{1- \psi}{2- \psi}\\$ 
Assuming the estimator is unbiased, M.O.M estimator satisfies $\\$
\begin{align}
\begin{aligned}
E[T] &= t \\
n\pi &= t \\
170\pi &= t \\
170 \times \frac{1- \psi}{2- \psi} &= t \\
170 \times \frac{1- \psi}{2- \psi} &= 8 \\
\frac{1- \psi}{2- \psi} = \frac{8}{170}\\
8(2- \psi) &= 170(1-\psi) \\
16 - 8\psi &= 170 - 170\psi \\
162 \psi &= 154 \\
\hat{\psi_0} &= \frac{154}{162} \approx 0.95 \\
\hat{\psi_0^{mom}} &= 0.95\\
\end{aligned}
\end{align}
(See Appendix 4 for the r code used for calculation) $\\$

To get the confidence intervals, the parametric bootstrap was used because we repeatedly draw new samples from a binomial distribution. 

We then used the Simple Percentile Method to find the confidence intervals because the $\hat{\psi_0^{mom}}$ we calculated which is 0.95 indicates that the sampling distribution would be skewed (See Appendix 5 for the r code used for bootstrapping and confidence interval). 

To find the P-value that tests $H_0: \psi_0 = 0.3$  $H_a: \psi_0 > 0.3$, we used 
empirical P-values where we calculated the p value by determining the proportion of time the test statistic value was at least as large as our observed value $t=8$ (See Appendix 6 for the r code used to calculate the P-value using empirical P-values). 
  
# Results: 
```{r psi_MLE, echo = FALSE, include = FALSE}
# Maximum Likelihood
loglik_binom <- function(psi, x, n){
  if(psi < 0 | psi > 1)
  return(NA)
  else {
  pi <- (1-psi)/(2-psi)
  return(log(choose(1000,50))+(x*log(pi))+((170-x)*log(1-pi)))
  }
}

MLE <- maxLik(logLik = loglik_binom, 
             start = 0.55,
             method = "NR", tol = 1e-4,
             x = 8, n = 170)
```  

```{r psi_MOM, echo = FALSE, include = FALSE}
# Method of Moments 
n <- 170
t <- 8
psi <- (2*t - n)/(-n+t)
```
To provide a comprehensive understanding of the data set and its relevance to our research questions, we first present the descriptive statistics and visualizations of the data, followed by inferential statistics.

## Observed Statistics
The table given in the introduction of the report summarizes the distribution of COVID-19 cases and the number of subjects in each group (BNT162b2 and Placebo), below is the data from the table in displayed in a segmented bar plot:

```{r segmented bar_plot, include = TRUE, echo = FALSE, fig.width = 6, fig.height = 3}
# make the data long for plot
data_long <- pivot_longer(Vaccine_data, cols = c("Covid19_Cases", "Non_Covid19_Cases"), 
                          names_to = "Status", values_to = "Count")

# segmented bar plot
ggplot(data_long, aes(fill = Status,
      y = Count, x = Group)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Covid19_Cases" = "red", "Non_Covid19_Cases" = "lightblue")) +
  labs(title = "Covid-19 Cases in Vaccine and Placebo Groups",
       x = "Group", y = "Number of Subjects", fill = "Status") +
  theme_minimal() 
```

The observed probability of contracting COVID-19 was `r vac_prob` , ($8 / 17411$) for the BNT162b2 group and was `r plac_prob` ($162 / 17511$) for the placebo group. The vaccine efficiency, calculated as the relative reduction in infection risk, was found to be `r round(vac_eff, 4)` \%. 

## Confidence Intervals: 
Using the methods described in the Statistical Methods portion of the report, we found the following Confidence Intervals:

**Maximum Likelihood Estimation (MLE):** Estimated efficacy ($\hat{\psi_0^{mle}}$): 0.95. \
The 95% Confidence interval is $[0.9145, 0.9855]$

**Method of Moments (MOM):** Estimated efficacy ($\hat{\psi_0^{mom}}$): 0.95. 
Bootstrap 95% Confidence interval is $[0.9097,	0.9817]$	

## Hypothesis Testing
Using the methods described in the Statistical Methods portion of the report, we found the following Hypothesis Test results:Null Hypothesis $H_0: \psi_0 = 0.3$ vs. Alternative Hypothesis $H_1: \psi_0 > 0.3$

**Likelihood Ratio Testing**
When $H_0$ is true, $W \approx \chi_1^2$ which calculated is approximately 0. This extremely small p-value indicates strong evidence against the null hypothesis, suggesting the vaccine efficacy is significantly greater than $30 \%$

**Empirical P-Value**
The Empirical P-value calculated using 1000 bootstrap samples under the null hypothesis, where the proportion of bootstrapped $\psi$ values exceeding the observed $\psi = 0.95$ was: p_val $= \frac{\sum (\psi^* \geq \psi_{\text{obs}})}{1000} \approx 0$

## Bootstrapped Sampling Distribution of \(\hat{\psi}\) & Sampling Distribution of \(\psi\) under the Null Hypothesis:
```{r bootstrapping, echo = FALSE, include = FALSE}
B = 1000
set.seed(1234)
boot_df_2 <- tibble(
  boot_psi = replicate(n = B, {
             n1 <- Vaccine_data$No_of_Subjects[1]
             n2 <- Vaccine_data$No_of_Subjects[2]
             
             BNT_covid <- Vaccine_data$Covid19_Cases[1]
             placebo_covid <- Vaccine_data$Covid19_Cases[2]
             
             BNT_resamp <- rbinom(1, n1, BNT_covid/n1)
             placebo_resamp <- rbinom(1, n2, placebo_covid/n2)
             
             pi_v <- BNT_resamp/n1
             pi_p <- placebo_resamp/n2
             pi <- pi_v/(pi_v+pi_p)
             
             boot_psi <- (1-2*pi)/(1-pi)
  })
)
#boot_df_2 %>% slice_head(n=5)
#boot_df_2 %>% summarise(boot_mean = mean(boot_psi), boot_sd = sd(boot_psi))
#boot_df_2 %>% summarise(lower = quantile(boot_psi, 0.025), upper = quantile(boot_psi, 0.975))
```

```{r boot_hist, echo = FALSE, warning = FALSE}
n <- nrow(boot_df_2)

# Sturges' formula
k_sturges <- ceiling(log2(n) + 1)

# calculate binwidth
binwidth <- (max(boot_df_2$boot_psi) - min(boot_df_2$boot_psi)) / k_sturges

# Plotting histogram with density curve overlay
hist1 <- ggplot(boot_df_2, aes(x = boot_psi)) +
  geom_histogram(aes(y = after_stat(count)), 
                 bins = k_sturges, fill = "lightpink", 
                 color = "hotpink") +
  geom_density(aes(y = after_stat(density) * B * binwidth), 
               color = "red", linewidth = 1) +
  labs(
    title = "Bootstrapped Sampling Distribution of",
    subtitle = expression(hat(psi) == frac(1 - 2 * pi, 1 - pi)),
    x = expression(hat(psi)),
    y = "Count"
  ) +
  theme_minimal()
```

```{r empirical pval, echo = FALSE}

B = 1000
set.seed(1234)

psi <- 0.3
pi_null <- (1 - psi)/(2 - psi)
total <- 34922

null <- tibble(
   psi_star = replicate(n = B, {
     covid = rbinom(1, size = total, prob = pi_null)
     pi = covid/total 
     psi = (1- 2*pi)/(1-pi)
     return(psi)
  })
)

hist2 <- ggplot(null, mapping = aes(x = psi_star)) +
  geom_histogram(aes(y = after_stat(count)),
                 bins = 30, fill = "lightblue",
                 color = "blue") +
  theme_minimal() +
  labs(
    title = "Sampling Distribution of Psi under the Null Hypothesis",
    subtitle = expression(psi == 0.3),
    x = "psi_star",
    y = "Count"
  ) +
  theme_minimal()
  

obs_t = 8
pval <- sum(null$psi_star >= obs_t)/ B
```

```{r, echo=FALSE, warning = FALSE, include = TRUE}
# Display hist1 and hist2 side by side
library(patchwork)
(hist1 | hist2)
```
The left histogram depicts the bootstrapped sampling distribution of the vaccine efficacy estimate derived from the Method of Moments (MOM) approach. The right histogram illustrates the sampling distribution of the vaccine efficacy estimate under the null hypothesis, where the efficacy is assumed to be 30%.

# Discussion/Conclusion

The analysis of the Pfizer vaccine data clearly indicates that the vaccine efficacy is significantly greater than the $30\%$ threshold set by the FDA for Emergency Use Authorization. The observed vaccine efficacy ($\hat{\psi}$) was estimated to be $95\%$, with both the Maximum Likelihood Estimation (MLE) and Method of Moments (MOM) approaches providing similar results.

\begin{itemize}
    \item The MLE estimate of vaccine efficacy was $95\%$, with a $95\%$ confidence interval ranging from 0.9144 to 0.9855. This interval does not include 0.3, indicating that the efficacy is significantly greater than $30\%$.
    \item The likelihood ratio test produced a test statistic of 121.6 with a p-value approximately equal to 0, strongly rejecting the null hypothesis that the vaccine efficacy is $30\%$.
    \item The MOM bootstrap approach yielded a confidence interval of (0.9097, 0.9817), further supporting the high efficacy estimate.
    \item The empirical p-value, calculated from the bootstrap samples under the null hypothesis, was effectively 0, reinforcing the rejection of the null hypothesis.
\end{itemize}

We used data from Polack et al.$^2$, and calculated an estimate for vaccination efficacy with methods different from the authors of the original paper. In the original paper the authors used a Bayesian beta binomial model with the transformation 
$\pi = \frac{\pi_v}{\pi_v+\pi_p}$ where $\pi_v$ and $\pi_p$ are the probabilities of an infection in the vaccine and placebo groups respectively. They estimated the vaccine efficacy $\psi = \frac{1-2 \pi}{1-\pi}$ to be 95\%. The 95\% Bayesian credible interval for $\psi$  was $[90.3\%, 97.6\%]$, and they reported that the Bayesian estimate for the probability that vaccine efficacy exceeded 30\% was larger than 0.9999. 

Our findings on the same data with differeing statistical methods corroborate the results presented in the paper, providing further support for the high efficacy of the Pfizer vaccine against COVID-19. The estimates we found in our analysis align with the 95% Bayesian estimate and the confidence intervals derived from both our MLE and MOM estimations are in line with the Bayesian credible interval reported in the paper. Additionally, our results affirm the conclusion drawn in the paper regarding the vaccine's effectiveness in surpassing the FDA's threshold of 
30% efficacy for Emergency Use Authorization. In summary, our study provides complementary evidence to the paper's findings, collectively contributing to the growing body of research affirming the considerable efficacy of the Pfizer vaccine in combating COVID-19.
  
# References

1. [COVID-19 Timeline](https://www.cdc.gov/museum/timeline/covid19.html#:~:text=March%2011%2C%202020,declares%20COVID%2D19%20a%20pandemic.)

2. [Polack et al.](https://www.nejm.org/doi/full/10.1056/nejmoa2034577)

3. [About the vaccine ](https://www.who.int/news-room/feature-stories/detail/who-can-take-the-pfizer-biontech-covid-19--vaccine-what-you-need-to-know)

4. [dashboard for covid cases](https://data.who.int/dashboards/covid19/deaths?n=o)

5. [About EUA](https://www.fda.gov/vaccines-blood-biologics/vaccines/emergency-use-authorization-vaccines-explained)

6. Stat 342 Materials: 
* 6.1 - Slide deck 24, slides 10,34 
* 6.2 - Slide deck 17, slide 5
* 6.3 - Assignment instructions
* 6.4 - Slide deck physician-health, slide 7
   
# Appendix
1. Maximum Likelihood estimation using maxLik
```{r psi_MLE1,echo = T, results = 'hide'}
loglik_binom <- function(psi, x, n){
  if(psi < 0 | psi > 1)
  return(NA)
  else {
  pi <- (1-psi)/(2-psi)
  return(log(choose(1000,50))+(x*log(pi))+((170-x)*log(1-pi)))
  }
}

MLE <- maxLik(logLik = loglik_binom, 
             start = 0.55,
             method = "NR", tol = 1e-4,
             x = 8, n = 170)
MLE
```
2. Using R as a calculator to calculate the SE and 95% confidence interval of  $\psi_0^{mle}$
```{r SE and CI1,echo = T, results = 'hide'}
SE <- sqrt(1/3045.805)
lower <- 0.95 - 1.96*SE
upper <- 0.95 + 1.96*SE
lower
upper
```
3. P-Value Calculation using Likelihood ratio test.
```{r pval1, echo = T, results = 'hide'}
pchisq(121.6, df = 1, lower.tail=F)
```
4. Using R as a calculator to calculate $\psi_0^{mom}$
```{r psi_MOM1, echo = T, results = 'hide'}
n <- 170
t <- 8
psi <- (2*t - n)/(-n+t)
psi
```
5. Bootstrap for confidence interval of $\psi_0^{mom}$
```{r bootstrapping1, echo = T, results = 'hide'}
B = 1000
set.seed(1234)
boot_df_2 <- tibble(
  boot_psi = replicate(n = B, {
             n1 <- Vaccine_data$No_of_Subjects[1]
             n2 <- Vaccine_data$No_of_Subjects[2]
             
             BNT_covid <- Vaccine_data$Covid19_Cases[1]
             placebo_covid <- Vaccine_data$Covid19_Cases[2]
             
             BNT_resamp <- rbinom(1, n1, BNT_covid/n1)
             placebo_resamp <- rbinom(1, n2, placebo_covid/n2)
             
             pi_v <- BNT_resamp/n1
             pi_p <- placebo_resamp/n2
             pi <- pi_v/(pi_v+pi_p)
             
             boot_psi <- (1-2*pi)/(1-pi)
  })
)

boot_df_2 %>% slice_head(n=5)
boot_df_2 %>% summarise(boot_mean = mean(boot_psi), boot_sd = sd(boot_psi))
boot_df_2 %>% summarise(lower = quantile(boot_psi, 0.025),
                      upper = quantile(boot_psi, 0.975))
```
6. Finding P-value for  $\psi_0^{mom}$ using Empirical P-values.
```{r empirical pval1, echo = T, results = 'hide'}
B = 1000
set.seed(1234)

psi <- 0.3
pi_null <- (1 - psi)/(2 - psi)
total <- 34922

null <- tibble(
   psi_star = replicate(n = B, {
     covid = rbinom(1, size = total, prob = pi_null)
     pi = covid/total 
     psi_star = (1- 2*pi)/(1-pi)
  })
)
  
obs_t = 8
pval <- sum(null$psi_star >= obs_t)/ B
```
7. Bar Plot of Covid-19 Cases in Vaccine and Placebo Groups
```{r segmented bar_plot1, echo = T, results = 'hide'}
# make the data long for plot
data_long <- pivot_longer(Vaccine_data, cols = c("Covid19_Cases", "Non_Covid19_Cases"), 
                          names_to = "Status", values_to = "Count")

# segmented bar plot
barplot <- ggplot(data_long, aes(fill = Status, y = Count, x = Group)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Covid19_Cases" = "red", "Non_Covid19_Cases" = "lightblue")) +
  labs(title = "Covid-19 Cases in Vaccine and Placebo Groups",
       x = "Group", y = "Number of Subjects", fill = "Status") +
  theme_minimal()
```
8. Histogram for Bootstrap Sampling Distribution of $\psi_0^{mom}$
```{r boot_hist1, echo = T, results = 'hide', warning=FALSE}
n <- nrow(boot_df_2)

# Sturges' formula
k_sturges <- ceiling(log2(n) + 1)

# calculate binwidth
binwidth <- (max(boot_df_2$boot_psi) - min(boot_df_2$boot_psi)) / k_sturges

# Plotting histogram with density curve overlay
hist1 <- ggplot(boot_df_2, aes(x = boot_psi)) +
  geom_histogram(aes(y = after_stat(count)), 
                 bins = k_sturges, fill = "lightpink", 
                 color = "hotpink") +
  geom_density(aes(y = after_stat(density) * B * binwidth), 
               color = "red", size = 1) +
  labs(
    title = "Bootstrapped Sampling Distribution of",
    subtitle = expression(hat(psi) == frac(1 - 2 * pi, 1 - pi)),
    x = expression(hat(psi)),
    y = "Count"
  ) +
  theme_minimal()
```
9. Empirical P Value Histogram
```{r empirical histogram, echo = T, results = 'hide'}

hist2 <- ggplot(null, mapping = aes(x = psi_star)) +
  geom_histogram(aes(y = after_stat(count)),
                 bins = 30, fill = "lightblue",
                 color = "blue") +
  theme_minimal() +
  labs(
    title = "Sampling Distribution of Psi under the Null Hypothesis",
    subtitle = expression(psi == 0.3),
    x = "psi_star",
    y = "Count"
  ) +
  theme_minimal()
```
